### 2022-09-29 Thesis/text/einleitung.tex
final v2

In theory, the robot can perform fully autonomous navigation, including mapping and localization. However, when something unexpected happens, the robot's autonomy can not be guaranteed. A default robot cannot represent all possible fail states and does not detect failures on a systematic level but only inside its navigation-related subsystems. Moreover, when failures inside these subsystems are discovered, the options to handle these problems are minimal and often do not deal with the problems in an autonomous way. The reliance on external problem-solving decreases the robot's usefulness when tasked with tangible goals. 
Therefore, this thesis centers around the research question of how behavior planning can increase the robustness and autonomy of a robot running ROS2. Behavior planning enables the robot to react to failures and problems of the system and can decide alternative courses of action to mitigate risks and finish the given tasks.

This thesis will explore possibilities for improving current systems by adding a dedicated behavior planning component. Furthermore, the thesis seeks to provide an implementation of exemplary behaviors and a system architecture to incorporate behavior planning in other robots, too.

@@

	\item Chapter \ref{cha:methoden} analyzes the current robotic systems and derives requirements from the key problem areas

	\item Chapter \ref{cha:implementierung} describes the implemented software approach including the overall system architecture

>>>>

In theory, the robot can perform fully autonomous navigation, including mapping and localization. However, the robot's autonomy can not be guaranteed when something unexpected happens. A default robot cannot represent all possible fail states and does not detect failures on a systematic level but only inside its navigation-related subsystems. Moreover, when failures inside these subsystems are discovered, the options to handle these problems are minimal and often do not deal with the problems in an autonomous way. The reliance on external problem-solving decreases the robot's usefulness when tasked with tangible goals. 
Therefore, this thesis centers around the research question of how behavior planning can increase the robustness and autonomy of a robot running ROS2. Behavior planning enables the robot to react to failures and problems of the system and can decide alternative courses of action to mitigate risks and finish the given tasks.

This thesis will explore possibilities for improving current systems by adding a dedicated behavior planning component. Furthermore, the thesis seeks to provide an implementation of exemplary behaviors and a system architecture to incorporate behavior planning in other robots.

@@

	\item Chapter \ref{cha:methoden} analyzes the current robotic systems and derives requirements from the critical problem areas

	\item Chapter \ref{cha:implementierung} describes the implemented software approach, including the overall system architecture

### 2022-09-29 Thesis/text/verwandt.tex
fertig?!
final v2

The Society of Automotive Engineers (SAE) defines vehicle autonomy based on the human driver's need for supervision and possible intervention when executing complex driving tasks \cite{J3016_202104}. Figure \ref{fig:sae_levels} depicts the different levels of autonomy of passenger cars on a scale between no automation to completely autonomous. The most significant distinction between the levels is between level two and three, because a system can take the complete control over the vehicle in certain conditions. The SAE levels zero to two use driver support functions with no real autonomy, while levels three to five employ automated driving features.
Another major level difference is located on the borderline between levels three and four as the vehicle can drive autonomously without a human operator who could control the vehicle as a safety fallback when the system demands it. A level four vehicle can theoretically be built without a steering wheel and pedals, because it is able to handle all situations it is deployed in without the need for manual driving. 
%The most significant distinction between the levels is between level three and four, because a vehicle can drive without a human driver as a safety fallback option.

@@

As shown in figure \ref{fig:autonomous_driving_architecture}, the sensory inputs, usually from multiple sensors, are processed to create an environment representation. Based on that representation, the system calculates how to get from its current position to the destination and creates a path to the destination. A second planning calculation is triggered to compute the motion commands while taking the system's constraints (e.g., size, turning radius) into account. In a final step, these motion commands then get converted to control the motors.

@@

Considering that with these architectures, vehicles can achieve an SAE level of autonomy up to level four and even five \cite{bacha2008odin}, a good starting point to increase autonomy in a robotic system would be to ensure that all of the functional modules in figure \ref{fig:autonomous_driving_architecture} are implemented and available. Based on this, higher autonomy levels are achievable by expanding the behavior planning module inside the planning layer.

@@

The behavior planning module contains a set of tailored behaviors to operate in a specific environment. The module's task is to always choose the best behavior to execute to ensure the best performance for a given task. In this sense, behavior is defined as a mapping of sensory inputs to a pattern of actions that carry out a specific task \cite{murphy2000}. A good behavior planning approach provides the robot with acceptable behaviors for every possible scenario the robot is operating in. Different scenarios pose different challenges for the behavior planning module, which has to decide on the best behavioral option to achieve a higher-level goal.

@@

During the emergence of behavior-based robots in the 1960s, the first type of behaviors implemented were simple reactive behaviors. This behavior type maps a sensory input directly to motor commands. In human behavior, this behavior resembles reflexes, like tapping on the kneecap, which unwillingly results in motion in the knee joint. This behavior pattern is not following the typical "Sense, Think, Act" loop, but shortcuts directly from sensing to acting \cite{desilva2008}. Reactive behaviors can be chained together to achieve more advanced and goal-driven behaviors.

Despite the possibility to carry out more complex robot behaviors, sequential behaviors remain largely on the level of reactive behaviors due to the lack of a planning and decision-making cycle during their execution. The computational load of reactive behaviors is low and therefore fast as no decision and planning process is taking place, which makes them suited in scenarios where real-time safety is of concern. This property makes this behavior type useful for system supervisors, where immediate reactions with low latency are sometimes required to ensure system safety.

@@

Systems purely based on reactive behaviors suffer the drawback that they do not allow for complex behaviors. In addition, creating sequences of reactive behaviors that produce the target behavior is a complicated task \cite{murphy2000}. Behaviors that involve planning and decision-making aspects are defined as deliberative. Unlike reactive behaviors, deliberative behaviors follow the "\textit{Sense - Think - Act}" paradigm, and they are not mapping sensory inputs directly into motor commands. Instead, deliberative-type behaviors can decide the best course of action before they act. Using deliberative behaviors inside of the planning module enables a robot to meet the definition of autonomy, especially in regards to the making of decisions and reacting to unforeseen circumstances.

A behavior planner with a deliberative approach can make decisions based on current sensor data and previously processed data. By incorporating older sensor information into the decision-making process, the deliberative behavior planner can predict changes in the environment. Using older data points allows the planner to proactively adapt the motion planning commands to better achieve goals in dynamic and uncertain environments.

@@

Concluding, the focus should be on creating deliberative behaviors to improve autonomy. Generally, these cognitively more complex behaviors mimic human skills and thus make a system more autonomous, because the need for a human operator can be decreased greatly.

%\textit{Talk about machine learning as deliberative behavior -> true autonomy bc learning and unknown scenarios, programmers can not cover every scenario, so a well designed/trained ai is needed to make a system really autonomous, but -> missing determinism and unaccountability of machines, hence we need behaviors that are still hard coded and deterministic, these can be highly automated. traditional behaviour and ai generated behaviours can and must coexist in the system (adler2019)}
%
%silva2008
%ingrand 2014
%vasiolopolous 2022

@@

Figure \ref{fig:hybrid_planning} shows the areas where reactive and deliberative behaviors have advantages and limitations. Deliberative planning does possess limitations during the execution of generated plans as the time horizon is long-term and not able to react to fast changes in the environment. When the system is not addressing these points, deliberative robots are very focused on a narrow problem domain and are not robust when the defined system boundaries are crossed\cite{arkin1998}.

@@

Figure \ref{fig:fsm} depicts an example state machine with different states and exemplary transitions between them. Inside of the states the letters indicate the existence of actions. $E$ stands for entry, $I$ for Input and $X$ for Exit. The arrows between the states illustrate the possible transition and transition conditions for every state. A state transition diagram is an excellent way to understand the system behavior, but can not show the details of the modeled behavior, like the actions \cite{wagner2006}.

@@

The execution time of finite state machines can be fast enough to control the motors of a bipedal robot to enable the robot to balance and walk despite unexpected height variations between steps \cite{park2013}. The high performance makes state machines very well suited for fast, reactive type behaviors while still allowing a deliberative and hierarchical approach to behavior planning.

Creating small state machines can be accomplished easily and quickly using various feature-rich and performant libraries \cite{fourakis2014}. Finite state machines are often used to model reactive and sequential behaviors comprehensively in autonomous driving functions \cite{ziegler2014}. These finite state machines use nested state machines inside another state machine. The nested state machines reduce the complexity of the state machine as the transitions do not need to be modeled because a nested state machine's start and final state are treated as just one state inside the bigger state machine.

Despite the possibility of nesting state machines, the effort to expand and maintain a FSM multiplies rapidly, when the system becomes increasingly larger to integrate more complex behaviors. The higher effort is due to the required creation of the transitions, transition conditions, and transition events. The FSM needs to be updated with each new state that gets introduced into the state machine as existing states need to be checked and corrected accordingly \cite{conner2017}. 
%
%allgeuer 2013
%connor 2017
%ziegler 2014

@@

\begin{table}[ht]
	\centering
	\caption{The five types of behavior tree nodes \cite{iovino2022}}
	\label{tab:node_types}
	\renewcommand{\arraystretch}{1.5}
	\resizebox{0.9\textwidth}{!}{%
		\begin{tabular}{ | m{0.13\textwidth} | m{0.13\textwidth}| m{0.2\textwidth} | m{0.2\textwidth} | m{0.2\textwidth} |} 
	 	\hline
	 	\textbf{Node type} & \textbf{Symbol} & \textbf{Succeeds} & \textbf{Fails} & \textbf{Running} \\ 
	 	\hline
	 	Sequence & -$>$ & If all children succeed & If one child fails & If one child returns running \\ 
	 	\hline
	 	Fallback & ? & If one child succeeds & If all children fail & If one child returns running \\ 
	 	\hline
	 	Parallel & -$>$ -$>$ & If $>=$ M children succeed & If $>$ N-M children fail & else \\
	 	\hline
	 	Action & shaded box & Upon completion & If impossible to complete & If node is still completing tasks \\
	 	\hline
	 	Condition & white oval & If true & If false & Never \\
	 	\hline
		\end{tabular}
	}
\end{table}

@@

On the other hand, the fallback node is an equivalent of a logical "Or" condition. The fallback node ticks its children nodes as long as they return "Failure" or "Running" and will return "Success" if one of the ticked nodes returned "Success". It will only return "Failure" if all children nodes returned "Failure". The primary node types and their control flow modification are listed in table \ref{tab:node_types}. Parallel control nodes are listed for completeness, but are not discussed further in this chapter due to their limited use in robotic applications.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/bt_example.jpg} 
	\caption{Example of a behavior tree \cite{iovino2022}}
	\label{fig:bt_example}
\end{figure}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/mdp.png} 
	\caption{Example of a fully observable Markov Decision Process \cite{andrew1999reinforcement}}
	\label{fig:mdp}
\end{figure}

@@

The difference between an MDP and a POMDP is that in a POMDP the system's current state can be unknown, and the system can make observations and estimate the state to a degree. The POMDP is defined as a tuple $<S, D, A, T, R, Z, O>$ in which the MDP tuple gets extended by \textit{Z} and \textit{O}. \textit{Z} is the observation, and \textit{O} is the probability function of seeing \textit{Z} in a given state \textit{S}.

POMDPs are well suited for applications in unknown environments because the system can make decisions if the perception information is imperfect, or the environment is not fully observable. The decision process is probabilistic and is more adaptable when confronted with unknown scenarios. This flexibility comes at the cost of determinism. The system's behavior is not guaranteed to be the same whenever it is subjected to the same sensory input \cite{krishnamurthy2016partially}.

>>>>

The Society of Automotive Engineers (SAE) defines vehicle autonomy based on the human driver's need for supervision and possible intervention when executing complex driving tasks \cite{J3016_202104}. Figure \ref{fig:sae_levels} on page \pageref{fig:sae_levels} depicts the different levels of autonomy of passenger cars on a scale between no automation to completely autonomous. The most significant distinction between the levels is between levels two and three because a system can take complete control over the vehicle in certain conditions. The SAE levels zero to two use driver support functions with no autonomy, while levels three to five employ automated driving features.
Another major level difference is located on the borderline between levels three and four as the vehicle can drive autonomously without a human operator who could control the vehicle as a safety fallback when the system demands it. A level four vehicle can theoretically be built without a steering wheel and pedals because it can handle all situations it is deployed in without requiring manual driving. 
%The most significant distinction between the levels is between levels three and four because a vehicle can drive without a human driver as a safety fallback option.

@@

As shown in figure \ref{fig:autonomous_driving_architecture} (p. \pageref{fig:autonomous_driving_architecture}), the sensory inputs, usually from multiple sensors, are processed to create an environment representation. Based on that representation, the system calculates how to get from its current position to the destination and creates a path to the destination. A second planning calculation is triggered to compute the motion commands while considering the system's constraints (e.g., size, turning radius). In a final step, these motion commands then get converted to control the motors.

@@

Considering that with these architectures, vehicles can achieve an SAE level of autonomy up to level four and even five \cite{bacha2008odin}, a good starting point to increase autonomy in a robotic system would be to ensure that all of the functional modules in figure \ref{fig:autonomous_driving_architecture} are implemented and available. Higher autonomy levels are achievable by expanding the behavior planning module inside the planning layer.

@@

The behavior planning module contains a set of tailored behaviors to operate in a specific environment. The module's task is always to choose the best behavior to execute to ensure the best performance for a given task. In this sense, behavior is defined as a mapping of sensory inputs to a pattern of actions that carry out a specific task \cite{murphy2000}. A good behavior planning approach provides the robot with acceptable behaviors for every possible scenario the robot is operating in. Different scenarios pose challenges for the behavior planning module, which has to decide on the best behavioral option to achieve a higher-level goal.

@@

During the emergence of behavior-based robots in the 1960s, the first type of behaviors implemented were simple reactive behaviors. This behavior type maps a sensory input directly to motor commands. In human behavior, this behavior resembles reflexes, like tapping on the kneecap, which unwillingly results in motion in the knee joint. This behavior pattern is not following the typical "Sense, Think, Act" loop but shortcuts directly from sensing to acting \cite{desilva2008}. Reactive behaviors can be chained together to achieve more advanced and goal-driven behaviors.

Despite the possibility of carrying out more complex robot behaviors, sequential behaviors remain primarily on the level of reactive behaviors due to the lack of a planning and decision-making cycle during their execution. The computational load of reactive behaviors is low and therefore fast as no decision and planning process is taking place, which makes them suited in scenarios where real-time safety is of concern. This property makes this behavior type useful for system supervisors, where immediate reactions with low latency are sometimes required to ensure system safety.

@@

Systems purely based on reactive behaviors suffer the drawback that they do not allow for complex behaviors. In addition, creating sequences of reactive behaviors that produce the target behavior is a complicated task \cite{murphy2000}. Behaviors that involve planning and decision-making aspects are defined as deliberative. Unlike reactive behaviors, deliberative behaviors follow the "\textit{Sense - Think - Act}" paradigm, and they are not mapping sensory inputs directly into motor commands. Instead, deliberative-type behaviors can decide the best course of action before they act. Using deliberative behaviors inside the planning module enables a robot to meet the definition of autonomy, especially in making decisions and reacting to unforeseen circumstances.

A behavior planner with a deliberative approach can make decisions based on current sensor data and previously processed data. The deliberative behavior planner can predict environmental changes by incorporating older sensor information into the decision-making process. Using older data points allows the planner to proactively adapt the motion planning commands to better achieve goals in dynamic and uncertain environments.

@@

Concluding, the focus should be on creating deliberative behaviors to improve autonomy. Generally, these cognitively more complex behaviors mimic human skills and thus make a system more autonomous because the need for a human operator can be significantly decreased.

%(adler2019)
%
%silva2008
%ingrand 2014
%vasiolopolous 2022

@@

Figure \ref{fig:hybrid_planning} (p. \pageref{fig:hybrid_planning}) shows the areas where reactive and deliberative behaviors have advantages and limitations. Deliberative planning does possess limitations during the execution of generated plans as the time horizon is long-term and not able to react to fast changes in the environment. When the system is not addressing these points, deliberative robots focus on a narrow problem domain and are not robust when the defined system boundaries are crossed\cite{arkin1998}.

@@

Figure \ref{fig:fsm} depicts an example state machine with different states and exemplary transitions between them. Inside of the states the letters indicate the existence of actions. $E$ stands for entry, $I$ for Input and $X$ for Exit. The arrows between the states illustrate the possible transition and transition conditions for every state. A state transition diagram is an excellent way to understand the system behavior, but it can not show the details of the modeled behavior, like the actions \cite{wagner2006}.

@@

The execution time of finite state machines can be fast enough to control the motors of a bipedal robot to enable the robot to balance and walk despite unexpected height variations between steps \cite{park2013}. The high performance makes state machines very well suited for fast, reactive type behaviors while allowing a deliberative and hierarchical approach to behavior planning.

Creating small state machines can be accomplished easily and quickly using various feature-rich, and performant libraries \cite{fourakis2014}. Finite state machines are often used to comprehensively model reactive and sequential behaviors in autonomous driving functions \cite{ziegler2014}. These finite state machines use nested state machines inside another state machine. The nested state machines reduce the complexity of the state machine as the transitions do not need to be modeled because a nested state machine's start and final state are treated as just one state inside the bigger state machine.

Despite the possibility of nesting state machines, the effort to expand and maintain a FSM multiplies rapidly when the system becomes increasingly larger to integrate more complex behaviors. The higher effort is due to the required creation of the transitions, transition conditions, and transition events. The FSM needs to be updated with each new state introduced into the state machine as existing states need to be checked and corrected accordingly \cite{conner2017}. 
%
%allgeuer 2013
%connor 2017
%ziegler 2014

@@

\begin{table}[ht]
	\centering
	\caption{The five types of behavior tree nodes \cite{iovino2022}}
	\label{tab:node_types}
	\renewcommand{\arraystretch}{1.5}
	\resizebox{0.99\textwidth}{!}{%
		\begin{tabular}{ | m{0.13\textwidth} | m{0.13\textwidth}| m{0.2\textwidth} | m{0.2\textwidth} | m{0.2\textwidth} |} 
			\hline
			\textbf{Node type} & \textbf{Symbol} & \textbf{Succeeds} & \textbf{Fails} & \textbf{Running} \\ 
			\hline
			Sequence & -$>$ & If all children succeed & If one child fails & If one child returns running \\ 
			\hline
			Fallback & ? & If one child succeeds & If all children fail & If one child returns running \\ 
			\hline
			Parallel & -$>$ -$>$ & If $>=$ M children succeed & If $>$ N-M children fail & else \\
			\hline
			Action & shaded box & Upon completion & If impossible to complete & If node is still completing tasks \\
			\hline
			Condition & white oval & If true & If false & Never \\
			\hline
		\end{tabular}
	}
\end{table}

@@

On the other hand, the fallback node is an equivalent of a logical "Or" condition. The fallback node ticks its children nodes as long as they return "Failure" or "Running" and will return "Success" if one of the ticked nodes returned "Success". It will only return "Failure" if all children nodes returned "Failure". The primary node types and their control flow modification are listed in table \ref{tab:node_types} (p. \pageref{tab:node_types}). Parallel control nodes are listed for completeness but are not discussed further in this chapter due to their limited use in robotic applications.

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.99\textwidth]{images/bt_example.jpg} 
	\caption{Example of a behavior tree \cite{iovino2022}}
	\label{fig:bt_example}
\end{figure}

@@

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.99\textwidth]{images/mdp.png} 
	\caption{Example of a fully observable Markov Decision Process \cite{andrew1999reinforcement, banerjee2021real}}
	\label{fig:mdp}
\end{figure}

@@

The difference between an MDP and a POMDP is that in a POMDP, the system's current state can be unknown, and the system can make observations and estimate the state to a degree. The POMDP is defined as a tuple $<S, D, A, T, R, Z, O>$ in which the MDP tuple gets extended by \textit{Z} and \textit{O}. \textit{Z} is the observation, and \textit{O} is the probability function of seeing \textit{Z} in a given state \textit{S}.

POMDPs are well suited for applications in unknown environments because the system can make decisions if the perception information is imperfect or the environment is not fully observable. The decision process is probabilistic and is more adaptable when confronted with unknown scenarios. This flexibility comes at the cost of determinism. The system's behavior is not guaranteed to be the same whenever it is subjected to the same sensory input \cite{krishnamurthy2016partially}.
